\section{Probability and Statistics}

\subsection{Probability Space}
Our intuitive notion of probability is related to uncertainty we have about events in the world, for example - will it rain tomorrow? How much will I get in the final exam? Although we cannot predict with certainty the answers to these questions, we usually know what is the set of all possible outcomes. This set is the first ingredient in any \emph{probability space}.
\begin{definition}
A \emph{sample space} $\Omega$ is a set that contains all possible outcomes. $\omega\in\Omega$ denotes a single outcome.
\end{definition}
Examples:
\begin{itemize}
\item Coin toss: $\Omega=\left\{ H,T\right\} $
\item Toss a coin until heads comes up: $\Omega=\left\{ T^{n-1}H:n\in\N\right\} $
\item Rolling  a die (singular of dice): $\Omega=\left\{ 1,\dots,6\right\}$
\item Rolling two dice: $\Omega=\left\{ 1,\dots,6\right\} ^{2}$
\item Waiting time at the post office: $\Omega=[0,\infty)$
\end{itemize}
Suppose someone threw a die and told us that the result is an even number. This information does not define unique element in $\Omega$, but it tells us that the outcome is an element in the subset $\left\{ 2,4,6\right\} $.
\begin{definition}
An \emph{event} $A$ is any subset of possible outcomes, $A\subseteq\Omega$.
\end{definition}
Quick reminder:
\begin{itemize}
\item $A^{c}$ (or sometimes $\overline{A}$) denotes the \textit{complement} of $A$, $A^{c}=\Omega\backslash A$ . In other words, $A^{c}$ is the event "$A$ did not occur".
\item We say that $A$ and $B$ are \textit{disjoint events} if $A\cap B=\emptyset$.
\item $\Omega$ and $\emptyset$ are also events. If $A$ and $B$ are two events, then $A\cup B$ is also an event, and so is $A\cap B$.
\end{itemize}

Intuitively, a probability function assigns to an event a number which quantifies our knowledge or belief about the possibility of observing any outcome in that event.

\subsection{Random Variables, discrete and continuous}

So far we have asked questions like: does an event happen or not and what is the probability of the event. But what about different questions, such as, "how much." For example, if we get a dollar if a coin flip turns head, we can ask how many dollars would we get after 3 rounds?  In order to answer this question, we should identify the sample space, which is
$\Omega=\{HHH,HHT,HTH,THH,HTT,THT,TTH,TTT\}$. Then we would need to define a reward function that will quantify ones's profit for each outcome of the experiment ($\omega\in \Omega$):
$X(HHH)=3,X(HHT)=2,$ and so on. This quantifying function $X$ is called a \emph{random variable} (RV).
\begin{definition}
	Given a probability space $\left(\Omega,\Dc\right)$, a real-valued \emph{random variable (RV)} is a function $X:\Omega\rightarrow\mathbb{R}$.
\end{definition}
\begin{remark}
	Any function $g\left(X\right)$ of a random variable is also a random
	variable (for instance $X^{2}$, $sgn \left(X\right)$, etc.).
\end{remark}
For a discrete $\Omega$ we define the \textbf{\textit{probability
		mass function (PMF)}} of $X$ (a \textit{discrete RV}) as
\[
\Dc\left(\left\{ X=x\right\} \right)=\sum_{\omega:X\left(\omega\right)=x}\Dc\left(\omega\right)
\]


Instead of writing $\Dc\left(\left\{ X=x\right\} \right)$  one sometimes writes $\Dc_X\left(\left\{ x\right\} \right)$ but more often,  simply $\Dc\left(x\right)$. This simplified notation, omitting any reference to the RV, is useful when the context is clear and we shall use it often. This omission is similar  to saying, for example,  "the probability of $(2,3)$ is 1/6" instead of  "the probability of getting $(2,3)$ by throwing the dice is 1/36".

\begin{example}
	A fair coin is tossed twice. The probability space is given by $\Omega=\left\{ T,H\right\} ^{2}$
	and  $\Dc\left(\omega\right)=\frac{1}{4}$ for any $\omega\in \Omega$. Let $X$ denote the number of heads obtained in each outcome ($\omega$). The possible values of $X$ are 0, 1, and 2,
	and
	\[
	\Dc\left(x\right)=\begin{cases}
		\frac{1}{4} & x=0,2\\
		\frac{1}{2} & x=1\\
		0 & \mbox{else}
	\end{cases}
	\]
	
\end{example}

For a continuous $\Omega$, we say that $X$ is a \textit{continuous
	RV} if there exists $f(x)\ge 0$ so that we can write, for every $D\subset\mathbb{R}$
\[
\Dc\left(X\in D\right)=\int_{D}f\left(x\right)dx
\]


In particular, this means that for any $a,b\in \mathbb{R}$, where $a\leq b$:
\[
\Dc\left(X\in\left[a,b\right]\right)=\int_{a}^{b}f\left(x\right)dx
\]

Again,  $f$ should actually be denoted by $f_X\left(x\right)$ to emphasize that it characterizes the RV $X$ but often the subscript $X$ is omitted.  In this course we assume that $f(x)$ is a regular function and therefore the probability mass function for a single point is 0: $\Dc\left(X=a\right)=\int_{a}^{a}f\left(x\right)dx=0$ for any $a\in\mathbb{R}$.

The function $f$ is called the \textbf{\textit{probability density
		function }}\textbf{(PDF)} of $X$. Note that -
\begin{itemize}
	\item $f\left(x\right)\ge0$
	\item $\int_{-\infty}^{\infty}f\left(x\right)dx=1$
	\item $f\left(x\right)$ is a probability \emph{density}, not a probability. For example, it could be that $f\left(x\right)>1$.\\
\end{itemize}

\subsection{Mean and Variance}
\begin{definition}
	The \textbf{\textit{expected value}} of a random variable
	$X$ is
	$$
	\E  \left[X\right]=\sum_{x\in\mathcal{X}}x\Dc\left(x\right)\qquad\mbox{or}\qquad \E \left[X\right]=\int_{-\infty}^{\infty}xf\left(x\right)dx
	$$
	
\end{definition}
Properties:
\begin{itemize}
	\item $\E \left[g\left(X\right)\right]=\sum g\left(x\right)\Dc\left(x\right)$,
	\item $\E \left[c\right]=c$
	\item Linear: $\E \left[aX+Y\right]=a\E \left[X\right]+\E \left[Y\right]$
	\item If $X$ and $Y$ are independent then $\E \left[XY\right]=\E \left[X\right]\E \left[Y\right]$
	since $\E [XY]=\sum\Dc(X=x,Y=y)x\cdot y = \sum_{x,y}\Dc(x)\Dc(y)xy=\E [X]\E [Y]$.
\end{itemize}

\begin{definition}
	Let $X$ be a RV with $\E \left[X\right]=\mu$, then the \textbf{\textit{variance}}
	of $X$ is
	\[
	V\left(X\right)=\E \left[\left(X-\mu\right)^{2}\right]
	\]
	The \textbf{\textit{standard deviation}} is defined as $\sigma=\sqrt{V\left(X\right)}$.
\end{definition}
Properties of the variance:
\begin{itemize}
	\item $V(X)\geq0$ , ~$V(X)=0\,\iff\,X$ is a constant.
	\item $V\left(X\right)=\E \left[X^{2}\right] - (\E \left[X\right])^{2}$
	\item $V(aX+b)=a^{2}V(X)$
	\item $V(X+Y)=\E \left[\left(X+Y-\E (X)-\E (Y)\right)^{2}\right]=V(X)+V(Y)+2\overbrace{\E \left[\left(X-\E [X]\right)\left(Y-\E [Y]\right)\right]}^{\sigma \left(X,Y \right)}$
	\item If $X_{1},\dots,X_{n}$ are independent then $V\left(\sum X_{i}\right)=\sum V\left(X_{i}\right)$\end{itemize}

\begin{definition}
	The \textbf{\textit{covariance}} of $X$ and $Y$ is denoted by $\sigma \left(X,Y \right)$ or $\sigma_{XY}$ and is defined as
\end{definition}
$$
\sigma(X,Y)\equiv \sigma_{XY}\equiv=\E \left[\left(X-\E [X]\right)\left(Y-\E [Y]\right)\right]=\E \left[XY\right]-\E [X]\E [Y]
$$

Properties:
\begin{itemize}
	\item $\sigma (X,Y)=\sigma \left(Y,X \right)$
	\item $\sigma \left(aX+b,cY+d \right) = a \cdot c \cdot \sigma \left(X,Y \right)$
	\item $\sigma(X,X)=V(X)$
\end{itemize}


\begin{example}
	Let $Z$ be a Bernoulli random variable. Calculate $Var[Z].$
\end{example}
Solution: $Var[Z] = \E [Z^2] - (\E [Z])^2 = p-p^2=p(1-p).$


\begin{example}
	Find the mean and variance of $X\sim U\left(\left[a,b\right]\right),$ where $U[a,b]$ is the  uniform distribution on $[a,b].$
	\begin{eqnarray*}
		\E \left[X\right] & = & \frac{1}{b-a}\int_{a}^{b}xdx=\frac{b+a}{2}\\
		\E \left[X^{2}\right] & = & =\frac{b^{2}+ab+a^{2}}{3}\\
		V\left(X\right) & = & \frac{b^{2}+ab+a^{2}}{3}-\frac{\left(b+a\right)^{2}}{4}=\frac{\left(b-a\right)^{2}}{12}
	\end{eqnarray*}
\end{example}

\subsection{A little Statistics: Mean and variance estimation}

This is a course in {\em Statistical} Machine Learning. So let's start with a little bit of statistics. When we study {\em Probability} we assume a known distribution and calculate probability of events of interest. When we study {\em Statistics} we turn the question on its head: using samples from an unknown probability distribution, we try to {\em estimate} or {\em test} properties of the unknown distribution.

Consider, for example, the mean and the variance.
The mean gives us a taste of around which point the data is distributed. The variance on the other hand gives us a taste of the concentration of the spread of the distribution around its mean.
Low variance usually means that the distribution is highly concentrated around the mean. High variance usually means the opposite.

In a probability course, given that we know the distribution, we learned how to calculate its mean and variance.
Now we would like to ask a different question. We are given {\em samples} that were drawn from an unknown distribution.
Can we {\em estimate}  the distribution's mean and variance using only those samples?


Formally, let $X$ be a random variable, let $\Dc \left(x\right)$ be its (unknown) distribution, and let $X_1,\ldots,X_n\overset{i.i.d}{\sim} D(x)$, where i.i.d  denotes \emph{independent and identically distributed} random variables. We are given \emph{data}
samples: $x_1,\ldots,x_n$, where $x_i$ is a {\em number} - the sample obtained by sampling $X_i$.
%The data has its own mean and variance which are likely to be, at least, slightly different than those of $\Dc \left(x\right),$ and therefore it is customary, in statistics, to use different names for these.
In statistics, $\ensuremath{\mathbb{E}}[X]$ is called the {\em population} mean - "population" added to emphasize that it is the mean of the unknown distribution, not the mean of the observed data.
Similarly, $V(X)$ is called the \textit{population variance}.

There are many ways to estimate $\ensuremath{\mathbb{E}}[X]$ and $V(X)$ based on those $n$ samples, and each one is called an \textit{estimator}. Throughout the course we are going to use the following estimators (commonly denoted with the "hat" symbol $\hat{\circ}$ to emphasize the fact that they are based on experimental observations and are not the real, namely, the population, mean and variance):
\begin{itemize}
	\item Sample mean (an estimator for the population mean)
	$$		\hat{\mu}_X = \frac{1}{n} \sum_{i=1}^n x_i$$
	\item Sample variance (an estimator for the population variance)
    $$		\hat{\sigma}_X^2 = \frac{1}{n-1} \sum_{i=1}^n (x_i-\hat{\mu}_X)^2$$
\end{itemize}

Whenever the context is clear, we will omit the subscript $X$ and write $\hat{\mu}$ and $\hat{\sigma}^2$ instead of $\hat{\mu}_X$ and $\hat{\sigma}_X^2$.
Note the $\frac{1}{n-1}$ factor. Why not use the observed variance of the samples,  $\frac{1}{n} \sum_{i=1}^n (x_i-\hat{\mu}_X)^2$, as the population variance estimator? \emph{Answer}: calculate the expected value of these estimators and you will find (perhaps surprisingly) that $\ensuremath{\mathbb{E}}[\hat{\sigma}_X^2]=V(X)$ while $\ensuremath{\mathbb{E}}[\frac{1}{n} \sum_{i=1}^n (x_i-\hat{\mu}_X)^2] = \frac{n-1}{n}V(X)$. So, in expectation, $\hat{\sigma}_X^2$ gets the population variance exactly. If the expected value of an estimator is equal to the parameter it is designed to estimate,  that estimator is said to be {\em unbiased}. Therefore, although $\hat{C}_X^2$ could have been considered, because of its form, a more natural way to estimate  $V(X)$, it is a \textit{biased} estimator. Biased estimators can be useful in certain cases, but for the mean and variance, we will stick with the unbiased ones.

