\chapter{Unsupervised Learning}
Up to this point our learning paradigm was of \textbf{supervised batch learning}. Given a sample space $\Xc$ and a label/response space $\Yc$, we were interested in \textbf{prediction}: finding some way to predict $y\in\Yc$ corresponding to a new unseen sample $x\in\Xc$, based on a training set of labeled samples $\trainset$. However, there are many learning problems that do not fit into this framework of supervised batch learning. In one such set of problems we still consider a domain set $\Xc$ but there is no label/response space. Namely, the data is of the form $S=\left\{\x_i\right\}^m_{i=1}$ where $\x_i\in\Xc$. Such problems are called \textbf{un-supervised} learning problems, and the goal is to infer different properties of $\Xc$. A few examples are:\\
\begin{itemize}
\item \textbf{Uncovering low-dimensional structures:} In some cases we have reason to believe that some given data, though represented in high dimensional, could actually be represented in a lower dimensional representation. Consider for example the MNIST database of handwritten digits. This coprus of 28-by-28 pixels grey-scaled images shows scans of people handwriting of the digits $0,\ldots,9$. Though each imange (sample) is represented in $\R^{28\times 28}$, there are very few variations on how people write digits. Therefore, it might be possible to \textbf{reduce the dimension} of the samples into a more compact, of lower dimension, representation.\\
\item \textbf{Clustering:} Often, when we are given a dataset, we are interested in grouping (or segmenting) it into subsets such that those samples within each cluster are in some way more "closely related" to each other than to samples in other subsets. We refer to these subsets as \textit{clusters}. Consider once more the MNIST digits dataset. In general, it seems logical that images of the same digit resemble each other more than images of different digits. If we would to try to cluster it (that is, segment its samples into different subsets) we would hope to get 10 separate subsets, each containing images of a specific digit.\\
\item \textbf{Anomaly detection:} Suppose we are interesting in detecting when some given system is not behaving "as usual". Consider an air conditioning system or power plant. We place sensors in the system and would like to get a warning when something is behaving "strange". We do not know what exactly a "strange" behavior would look like, but it might be caused due to some mechanical malfunction, a software problem or even a cyber attack. We monitor the state of the system at all times. If we have installed $d$ sensors, each time we take a reading of the system we get a sample $\x_i\in\R^d$. We train out learning system on the readings $\x_1,\ldots,\x_m\in\R^d$ where we believe these represent the normal state of the system. Now, given a new reading $\x\in\R^d$, is it "normal"? or is there something wrong? We are required to make a decision without have seeing the system in the "wrong"/"abnormal"/"strange" state before.
\end{itemize}

\section{Dimensionality Reduction}
Dimensionality reduction is the process of mapping some high dimensional space (the ambient space) to some new low dimensional space (the intrinsic space). Given a dataset $\x_1,\ldots,\x_m\in\R^d$ find some mapping $f:\R^d\rightarrow\R^k$ for $k<d$, where $f\left(\x_1\right),\ldots,f\left(\x_m\right)$ still resembles $\x_1,\ldots,\x_m$ in some sense. Different dimensionality reduction techniques are usually separated to linear- and non-linear techniques depending on the selected $f$. 
\\~\\
There are different reasons to study and apply dimension reduction:
\begin{itemize}
	\item \textbf{Learnability:} Some learning algorithms work better when the dimension $d$ is small compared with the training sample size $m$, and might even fail if $d$ is too large. Recall for example the case of linear regression. Though in both cases $m<d$ and $d<m$ we have a closed form solution, we have seen that if $d<m$ the results are numerically unstable. \\
	\item \textbf{Computation:} For many algorithm the time- and space complexity depends on the dimension $d$. By reducing the data dimensionality we can use fewer computational resources to perform the learning task we intended. \\
	\item \textbf{Visualization:}  Visualization is an extremely useful tool both for explaining and exploring our data. If we can reduce dimension to $k\leq 4$ then we can plot the data (possibly on a 3 dimensional axis, with the forth dimension represented by color).
\end{itemize}


\subsection{Principal Component Analysis (PCA)}
Let $\x_1,\ldots,\x_m\in\R^d$ be some dataset and suppose that it approximately resides in some lower $k$-dimensional linear subspace of $\R^d$. We would like to find some linear transformation of the data to project it on that lower dimension subspace. To do so we should solve to challenges:
\begin{itemize}
	\item There are infinitely many subspaces we could consider. Which subspace should we choose and how should we acquire it?
	\item Even if we knew the subspace, projecting the data-points $\x\in\R^d$ onto the subspace does not yet rerude the dimension. We would want to find a way to represent each sample by the $k$ coordinates of that sample \textit{in} the subspace. This is also known as the embedding of the samples.
\end{itemize}

In the case of Principal Component Analysis (PCA) given a dataset $\x_1,\ldots,\x_m\in\R^d$ we are searching for some linear transformation such that we minimize the squared error between the original samples and their transformation:
\begin{equation}
f^*=\underset{f}{argmin} \sum^m_{i=1}\norm{\x_i - f\left(\x_i\right)}^2
\end{equation}
This problem can be formulated in four different ways:
\begin{itemize}
	\item Finding the closest affine subspace to the points.
	\item Finding the affine subspace that retains most of the variation in the data
	\item Finding the affine subspace that minimizes the distortion of distances between points in the ambient compared to the intrinsic spaces.
	\item Generalization of linear regression
\end{itemize}~\\
Here, we will discuss the first two, most common, formulations.

\subsubsection{Closest Affine Subspace}
To simply the problem of finding the closest affine subspace, let us assume that the data is centered around the origin and as such instead of searching for some affine subspace, we are searching for an "ordinary" subspace. Therefore, we are searching for some linear map $W:\R^d\rightarrow\R^k$ and an "inverse" linear map $U:\R^k\rightarrow\R^d$
\begin{equation}
W^*,U^* = \underset{W\in\R^{d\times k},U\in\R^{k\times d}}{argmin} \sum_{i=1}^m \norm{\x_i-UW\x_i}^2
\label{eqn:pca1}
\end{equation}


\begin{lemma}
Let $\left(U,W\right)$ be a solution to (\ref{eqn:pca1}). Then $U$'s columns are orthonormal and $W=U^\top$.
\end{lemma}
\begin{proof}
Let $U,W$ be some matrices and consider the mapping $\x\mapsto \left(UW\right)\x$. The matrix $\left(UW\right)$ is a $d$-by-$d$ matrix of rank $k$, whos image is a $k$ dimensional subspace of $\R^d$. Denote $S\coloneqq Im\left(UW\right)$. As $\x\in\R^d$ and $\left(UW\right)\x\in S$, it holds that the projection that minimizes $\norm{\x-UW\x}_2$ is the orthogonal projection onto the subspace. In addition, the point in $S$ closest to $\x$, namely the orgothonal projection pf $\x$ onto $S$, is given by $VV^\top\x$ where the columns of $V$ are an orthonormal basis of $S$:
$$ \forall \vv{u}\in S\quad \norm{\x-\vv{u}}_2 \geq \norm{\x-VV^\top\x}_2 $$
Thus, the solution to \ref{eqn:pca1} is $U$ with orthonormal columns and $W\coloneqq U^\top$. 
\end{proof}

Based on the above lemma, we can now write an equivalent problem to the PCA problem presented in \ref{eqn:pca1}. Observe that:
$$
\begin{array}{ccl}
\norm{\x-UU^\top\x}^2 & = & \norm{\x}^2 -2\x^\top UU^\top\x + \x^\top UU^\top UU^\top\x \\
& = & \norm{\x}^2 -\x^\top UU^\top\x\\
& \overset{\left(*\right)}{=} & \norm{\x}^2 -\left(U^\top\x\right)^\top U^\top\x\\
& = & \norm{\x}^2 - trace\left(U^\top\x \left(U^\top\x\right)^\top\right) \\
& = & \norm{\x}^2 - trace\left(U^\top\x\x^\top U\right)
\end{array}$$

where $\left(*\right)$ is because $\forall \vv{v},\vv{u}\in\R^k \,\, trace\left(\vv{u}\vv{v}^\top\right)=\vv{v}^\top\vv{u}$. As the trace is a linear operator we can re-write an equivalent problem:
\begin{equation}
U^* = \underset{U\in\R^{d\times k},U^\top U=I}{argmax} \sum_{i=1}^m trace\left(U^\top\x_i\x_i^\top U\right) = \underset{U\in\R^{d\times k},U^\top U=I}{argmax} trace\left(U^\top\sum_{i=1}^m \x_i\x_i^\top U\right)
\label{eqn:pca2}
\end{equation}



\begin{theorem}
Let $A=\sum\x_i\x_i^\top$ and let $\vv{u}_1,\ldots,\vv{u}_k$ be the $k$ leading eigenvectors of $A$. Then, the solution to the PCA problem is given by $U\in\R^{d\times k}$ whose columns are $\vv{u}_1,\ldots,\vv{u}_k$.
\end{theorem}
\begin{proof}
Denote $A=\sum\x_i\x_i^\top$. Then, we need to solve $$ \underset{U\in\R^{d\times k},U^\top U=I}{argmax} trace\left(U^\top A U\right)$$
Notice that $A$ is a square symmetric matrix so let $A=VDV^\top$ be the EVD of $A$, where the diagonal of $D$ are the eigenvalues of $A$ in decreasing order $\lambda_1\geq \ldots \geq \lambda_d$ and the columns of $V$ are the corresponding eigenvectors. So:
$$
\begin{array}{cclcl}
trace\left(U^\top AU\right) & = & trace\left(U^\top V D V^\top U\right) & \overset{\left(*\right)}{=} & trace\left(B^\top D B\right)\\
& = & trace\left(DBB^\top\right) &=& \sum_{j=1}^d\left[DBB^\top\right]_{jj} \\
& = & \sum_{j=1}^d \left[D\right]_{j} \left[BB^\top\right]_{\cdot,j} & = & \sum_{j=1}^d \lambda_j\cdot \left[BB^\top\right]_{jj} \\
& = & \sum_{j=1}^d \lambda_j\cdot \sum_{i=1}^k B_{ji}^2
\end{array}
$$
where $B=V^\top U\in \R^{d\times k}$. Notice that 
$$
\begin{array}{c}
B^\top B=U^\top VV^\top U = I_d\\
\Downarrow\\
\indc{j\neq i} = \inprod{\left[B^\top\right]_j}{\left[B\right]_{\cdot,i}} = \inprod{\left[B\right]_{\cdot,j}}{\left[B\right]_{\cdot,i}}
\end{array}$$
meaning that the columns of $B$ are orthonormal, and therefore $\sum_{j=1}^d\sum_{i=1}^k B_{ji}^2 = k$. Based on $B$, define $\widetilde{B}=\left[\begin{array}{c|c} B & M\end{array}\right] \in \R^{d \times d}$ with $M\in\R^{d\times \left(d-k\right)}$ completing an orthonormal basis in $R^d$. If so, then $\forall j\,\, \sum_{i=1}^d\widetilde{B}^2_{ji}=1$, which implies that $\sum_{i=1}^k B_{ji}^2$. As such 
$$ trace\left(U^\top A U\right) = \sum_{j=1}^d \lambda_j\cdot \sum_{i=1}^k B_{ji}^2 \leq \underset{\beta\in \left[0,1\right]^d, \norm{\beta}_1\leq k}{max} \sum_{j=1}^d \lambda_j \beta_j =\sum_{j=1}^k \lambda_k$$

To conclude the proof, let $U$ be the matrix whose columns are the $k$ leading eigenvalues of $A$ then $$
U^{\top}AU=\left[\begin{array}{c}
u_{1}^{\top}\\
\vdots\\
u_{k}^{\top}
\end{array}\right]A\left[\begin{array}{ccc}
u_{1} & \cdots & u_{k}\end{array}\right]=\left[\begin{array}{c}
u_{1}^{\top}\\
\vdots\\
u_{k}^{\top}
\end{array}\right]\left[\lambda_{1}\begin{array}{ccc}
u_{1} & \cdots & \lambda_{k}u_{k}\end{array}\right]=diag\left(\lambda_{1},\ldots,\lambda_{k},\overset{d-k}{\overbrace{0,\ldots,0}}\right)
$$
and $trace\left(U^\top A U\right) = trace\left(diag\left(\lambda_{1},\ldots,\lambda_{k},0,\ldots,0\right)\right)=\sum^k \lambda_i$, as requested.
\end{proof}

~\\\paragraph{Generalizing for affine subspaces:} In the above theorem we in-fact found the closest subspace- and not the closest affine subspace to the data. To find the closest affine subspace, we would like to allow the mapping $W$ to be \textbf{affine}. To achieve this we generalize the above by considering $W:\R^d\rightarrow\R^k$ to be of the form $$ W\left(\x\right)\coloneqq\widetilde{W}\left(\x-\mu\right),\quad \mu\in\R^d,\, \widetilde{W}:\R^d\rightarrow\R^k$$ 
This allows us to ``shift`` the data before applying the linear map $\widetilde{W}$. When adding $\mu$ to the optimization problem above, we find that the minimizer $\mu$ is given by: $ \mu = \frac{1}{m}\sum \x_i $. This is the empirical mean of the data, often denoted by $\overline{\x}$. So, in order to find the closest affine subspace to the data we center the matrix $A$: $$ A\coloneqq \sum_{i=1}^m \left(\x_i -\overline{\x}\right)\left(\x_i -\overline{\x}\right)^\top $$

yielding the following pseudo-code for the PCA algorithm:

\begin{algorithm}
	\caption{PCA}\label{pca_pseudo}
	\begin{algorithmic}
		\Procedure{PCA}{$\X$, $k$}\Comment{$\X$ The design matrix of $m$ samples and $d$ features}
		\State Compute $A\gets \sum_{i=1}^m \left(\x_i-\overline{\x}\right)\left(\x_i-\overline{\x}\right)^\top$ 
		\State Let $\vv{u}_1,\ldots,\vv{u}_k$ be the eigenvectors of $A$ corresponding the largest eigenvalues
		\State \textbf{return} $\vv{u}_1,\ldots,\vv{u}_k$
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

~\\ The eigenvalues of $A$, $\lambda_1\geq\ldots\geq\lambda_d$ are referred to as the \textbf{Principal Values} of $\X$. The eigenvectors of $A$, $\vv{u}_1,\ldots,\vv{u}_d$ are referred to as the \textbf{Principal Vectors}. Notice that the matrix $A$ is a $d$-by-$d$ positive semi-definite matrix. Therefore
\subsubsection{Maximum Retained Variance}

\todo{show that minimizes var}
\todo{geometric view}
\todo{explained variance}



\subsubsection{Projection- vs. Coordinates of Data-Points}
\subsubsection{Variants of PCA}
\subsection{Euclidean Embedding}

\section{Clustering}
A very useful set of learning problems is of clustering. Often, either as part of data exploration or as part of the main analysis, we are interested in partitioning our data into \textbf{meaningful} groups. For example, given a corpus of images we might want to divide them into different groups such as nature/urban/people/etc. photographs; or, given the set of genes in the human genome, we might want to group together genes associated with specific diseases. In both these examples we are only given the samples (images or genes) but we do not have any \textbf{ground truth} (labels). We are not given the information of what is seen in each image, nor for each disease what genes are associated with it. 

~\\We would therefore want to define some measure of \textit{similarity} between samples of a given domain. Using this similarity we could split the data into different subsets, where intuitively samples within a given set are \textit{more similar} to one another compared to samples of different sets.

\begin{remark}
This form of clustering, where we partition the data into distinct non-overlapping groups is also referred to as ``hard assignment`` as each sample is assigned to one specific subset. Sometimes, rather than assigning to some specific cluster we are interested in ``partly assigning`` to multiple clusters. This is referred to as ``soft assignment/clustering``.
\end{remark}

\subsection{K-Means}
Though there are many different approaches to perform clustering, common to many is the notion of defining some representative data-point of the cluster. Then, clustering of data-points give the given sample is perform with respect to these cluster representatives.

\begin{definition}
A partition of a dataset $\left\{\x_i\right\}_{i=1}^m$ is a set $C_1,\ldots,C_k$ such that $\left\{\x_i\right\}_{i=1}^m = \biguplus_{j=1}^k C_j$
\end{definition}

Given some partition over the data and the representative data-points, we can define a \textbf{cost function} for the partition. For some metric (i.e. distance function) over the data $d:\Xc\times\Xc\rightarrow \R_+$ we define:
\begin{equation}
G_d\left(C_1,\ldots,C_k, \mu_1,\ldots,\mu_k\right) \coloneqq \sum_{j=1}^k \sum_{\x\in C_j} d\left(\x,\mu_j\right)
\end{equation}
where $\mu_1,\ldots,\mu_k$ are the representatives of clusters $C_1,\ldots,C_k$. Using this function, the goal is to find the partitioning that minimizes the following:
\begin{equation}
\left\{C_1,\ldots,C_k\right\}^* = \underset{\left\{C_j,\mu_j\right\}_{j=1}^k}{argmin} G_d\left(C_1,\ldots,C_k,\mu_1,\mu_k\right) = \underset{\left\{C_j,\mu_j\right\}_{j=1}^k}{argmin} \sum_{j=1}^k \sum_{\x\in C_j} d\left(\x,\mu_j\right)
\label{eqn:cluster_cost}
\end{equation}

~\\In the case of the $k$-means algorithm the cluster representatives are chosen to be: $$ \mu_j\left(C_j\right) \coloneqq \underset{\mu\in\Xc}{argmin}\sum_{\x\in C_j} d\left(\x,\mu\right)  $$

In order to find the minimizers of \ref{eqn:cluster_cost} we would have to navigate the space of all possible partitions of $m$ objects into $k$ subsets. As there are an exponential amount of such subsets, minimizing the cost function $G$ is NP-hard, and we must resort to \textbf{heuristics}. The most famous heuristic for minimizing $G$, when $d$ is the Euclidean distance, is k-means.

\begin{algorithm}
	\caption{K-Means}\label{kmeans_pseudo}
	\begin{algorithmic}
		\Procedure{K-Means}{$\X$, $k$}\Comment{$\X$ The design matrix of $m$ samples and $d$ features}
		\State Choose initial centroids $\mu_1,\ldots,\mu_k$.
		\While{Not converged}
			\State \textbf{Assignment:} Assign each point $\x_i$ to the centroid closest to it: $$ C_j^{\left(t\right)}\coloneqq \left\{\x | \mu_j = \underset{\mu}{argmin}\,d\left(\x,\mu\right),\x\in\X\right\} $$
			\State \textbf{Update}: Adjust cluster centroids by: $ \forall j\in\left[k\right] \quad \mu_j^{\left(t+1\right)} \coloneqq \left|C_j^{\left(t\right)}\right|^{-1}\sum_{\x\in C_j^{\left(t\right)}}\x $
		\EndWhile
		\State \textbf{return} $C_1,\ldots,C_k$
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

~\\ This clustering approach uses \textbf{Lloyd's Algorithm} to minimize $G$ using an iterative strategy where each time we \textbf{alternate} between minimizing two terms. We first define a partition of the dataset, associating points to subsets by their nearest centroid. These subsets are called Voronoi cells. Then we re-calculate the cluster's centroid.

~\\\todo{finish - show example of run (animation)\\discuss suboptimal solutions and multiple optimal solutions. Discuss outliears\\ Choosing $k$ (and recall elbow plot in linear regression)}

